The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-06-20 paper.qmd
started signal supression restructure

# What is uncertainty?
In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.  Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

## Definitions of Uncertainty in the Literature
Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty".

While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects. The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. The difficulties authors have with designing a clear definition of uncertainty likely comes from the fact that these definition fail to identify precicely what it is that *creates* uncertainty. These authors have failled to identify that uncertainty is simply a by product of inference.

## Defining uncertainty with respect to inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

@@

It is clear that uncertainties dependence on inference has gone unnoticed in the literature because authors often mention the by products of this fact as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017]. With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.

The belief that uncertainty is a feature of the data, not the variable of interest, leads to several flabbergasting experiments in the uncertainty visualisation space. In these experiments, it is unclear how the participants were supposed to relate what they were shown to the uncertainty distribution the authors had in mind (or if the authors had any distribution in mind at all). For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. In another experiment, @Hofman2020 asked participants to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.

@@

**UP TO HERE FOR RESTRUCTURE/REWRITE**

>>>>

# The purpose of uncertainty visualisation
## Other motivations 
## Signal supression

# When is signal supression appropriate?
Signal suppression is not always possible in a visualisation. Whether or not uncertainty should be used for signal-suppression largely depends on the purpose of the visualisation and the definition of uncertainty.

## What is uncertainty
### Current definitions of uncertainty
Most definitions of uncertainty treat it as a feature of the data, as something that exists in our data no matter it's purpose. This is not entirely accurate.

This view of uncertainty leads to a wide range of definitions that are designed to be encompassing, but struggle to capture the full picture. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Some authors use taxonomies to split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.

The constantly shifting definitions make it hard to pin down and quantify uncertainty. The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

The constanly shifting definitions might be a result of the view of uncertainty as a latent variable that can exist without an explicit purpose. This is not true, as uncertainty only exists as a by-product of inference.

### The case for uncertainty as a by product of inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

@@

Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Whether or not these elements are relevant is highly dependent on what statistic you are trying to draw inference on, and by extension, the purpose of your visualisation.

## EDA vs Communication
## If signal supression is not the goal

# How should be test uncertainty visualisations?
## Uncertainty as signal
## Uncertainty as noise
## Other uncertainty experiments 
-----------------------------------------------

# UP TO HERE FOR SORTING

# What is uncertainty?

## Defining uncertainty with respect to inference

For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. In another experiment,

@Hofman2020 asked participants to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.

@@

-----------------------------------------------

# UP TO HERE FOR RESTRUCTURE/REWRITE
